Convective heat transport in stratified atmospheres at low and high Mach number

### 2016-10-26 tex/nusselt_paper.tex
start adding Jeff's notes

%%%%% Body of the paper
\section{Introduction}
\refstepcounter{section}
\label{sec:intro}
Convection is essential to heat transport in the cores of high mass stars, the
envelopes of low mass stars, and the atmospheres of terrestrial and jovian planets. In such systems, convection
occurs in the presence of the atmospheric stratification which can be small but extends up to 
14 density scale heights in the Sun's convective envelope.
A basic understanding of the
properties of compressible convection in stratified media is important to understanding systems in astrophysics
and planetary sciences.  Numerical constraints have often restricted studies of stratified convection to
moderately high Mach numbers, appropriate to regions near the Sun's surface.  As such, we know few
fundamental properties of the low-Mach number stratified convection which occurs in the deep solar interior.

Early numerical experiments on stratified convection
in two \cite{graham1975, chan&all1982,
hurlburt&all1984, cattaneo&all1990} and three \cite{cattaneo&all1991, brummell&all1996} dimensions
revealed a number of basic properties in the moderate-to-high Mach number regime.
In the widely-studied \RB (hereafter RB) problem, upflows and downflows are symmetrical and
the temperature gradient approaches zero in the convective interior causing the conductive flux to similarly 
disappear.  Highly stratified convection exhibits narrow downflow lanes and broad upflow regions.
Furthermore, the \emph{entropy} gradient is negated by convection rather than the temperature gradient, such
that in the presence of perfectly efficient convection a significant component of the flux is still
carried by conduction.

In RB convection, there exist two primary control parameters: the Rayleigh number (Ra), the ratio of
buoyant driving to diffusive damping, and the Prandtl number (Pr), the ratio of viscous to thermal
diffusivity.  These numbers couple with the aspect ratio of the physical domain and the boundary conditions
to entirely control the
dynamics of the convection.  In stratified atmospheres, in addition to specifying the equation of state and
fundamental properties of the gas, the two control parameters of RB convection are joined by the degree of
stratification across the domain and the characteristic Mach number (Ma) of the convective flows.  
Polytropically stratified atmospheres, such as those used in early studies, are an ideal extension of
RB convection into the stratified realm as the two additional control parameters are directly linked to
basic properties of the atmosphere.  The density stratification is set by the number of density scale heights
the atmosphere spans (\nrho), and Ma is controlled by the superadiabatic excess ($\epsilon$),
the deviation of the polytropic index from the adiabatic polytropic index \cite{graham1975}.

@@

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
In order to compare our results with previous studies and in an effort to examine a simplest case,
we study a fluid composed of monatomic ideal gas particles with an adiabatic index of $\gamma = 5/3$ and
whose equation of state is $P = R\rho T$. 
The initial stratification is polytropic and the gravitational
acceleration and conductive heat flux are invariant throughout the depth of the atmosphere. We specify that both 
the thermal conductivity, $\kappa$, and the temperature gradient, $\grad T_0$, are constant while constructing
our atmosphere such that $\bm{F}_{\text{cond,0}} = -\kappa \grad T_0 = \text{constant}$.
Under these assumptions, solving the equation of hydrostatic equilibrium produces an atmosphere defined by
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{00}(z_0 - z)^m \\
T_0(z)    &= T_{00}(z_0 - z).
\label{eqn:polytrope}
\end{split}
\end{equation}
Thermodynamic variables are nondimensionalized at the top of the atmosphere as 
$P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, requiring $z_0 \equiv L_z + 1$ and $R = T_{00} = \rho_{00} = 1$.
By this choice, the non-dimensional length scale is the inverse temperature gradient scale and
the timescale is the isothermal sound crossing time of this unit length.
$z$ increases upwards within the bounds $z =[0, L_{z}]$, where $L_{z} = e^{n_{\rho}/m} - 1$ is
determined by the number of density scale heights the atmosphere spans, $n_\rho$.
Throughout this letter, we set $n_{\rho} = 3$ such that the density varies by a factor of 20.
The polytropic index is such that $m = m_{ad} - \epsilon$
where $m_{ad} \equiv (\gamma-1)^{-1}$ is the adiabatic polytropic index and $\epsilon$ is the
superadiabatic excess.
The subsequent entropy gradient at the top of the atmosphere is $\grad S(L_z) = -\epsilon$.  
The characteristic timescale of convective dynamics
is related to the atmospheric buoyancy time, $t_{\text{b}} = \sqrt{L_z/g\epsilon}$, with $g = (m+1)$.
We will utilize buoyancy time units throughout this letter.

>>>>

%%%%% Body of the paper
\section{Introduction}
\refstepcounter{section}
\label{sec:intro}
Convection is essential to heat transport in the cores of high mass stars,
the envelopes of low mass stars, and the atmospheres of terrestrial and 
jovian planets.  The convective dynamics are influenced by the atmospheric 
stratification, which is small in some systems but extends up to 
14 density scale heights in the Sun's convective envelope.
Understanding the fundamental
properties of compressible convection in stratified media is essential
to characterizing systems in astrophysics and planetary sciences.  
Numerical constraints have often restricted studies of compressible
convection to moderately high Mach number (Ma), appropriate to regions near 
the Sun's surface.  Few fundamental properties of the 
low-Ma stratified convection which occurs in the deep solar interior are known.

Early numerical experiments on stratified convection
in two \cite{graham1975, chan&all1982,
hurlburt&all1984, cattaneo&all1990} and three 
\cite{cattaneo&all1991, brummell&all1996} dimensions
revealed a number of basic properties in the moderate-to-high 
Ma regime. In the widely-studied \RB (hereafter RB) problem, 
upflows and downflows are symmetrical and
the temperature gradient approaches zero in the convective interior 
causing the conductive flux to similarly 
disappear.  Highly stratified convection exhibits narrow downflow 
lanes and broad upflow regions.
Furthermore, the \emph{entropy} gradient is negated by convection 
rather than the temperature gradient, such
that in the presence of perfectly efficient convection a significant 
component of the flux is still carried by conduction.

In RB convection, there exist two primary control parameters: 
the Rayleigh number (Ra, the ratio of
buoyant driving to diffusive damping) and the Prandtl number 
(Pr, the ratio of viscous to thermal
diffusivity).  These numbers coupled with the aspect ratio of 
the physical domain and the boundary conditions
determine the dynamics of the convection.  In stratified atmospheres, 
in addition to specifying the equation of state and
fundamental properties of the gas, the two control parameters of 
RB convection are joined by the degree of
stratification across the domain and the characteristic 
Ma of the convective flows.  
Polytropically stratified atmospheres, such as those used in 
early studies, are an ideal extension of
RB convection into the stratified realm as the two additional 
control parameters are directly linked to
basic properties of the atmosphere.  The density stratification is 
set by the number of density scale heights
the atmosphere spans (\nrho), and Ma is controlled 
by the superadiabatic excess ($\epsilon$,
the deviation of the polytropic index from the adiabatic polytropic 
index \cite{graham1975}).

@@

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
In order to compare our results with previous studies and in an effort to examine a simplest case,
we study a fluid composed of monatomic ideal gas particles with an adiabatic index of $\gamma = 5/3$ and
whose equation of state is $P = R\rho T$. In constructing an initial atmosphere, we assume
that the gravitational
acceleration and conductive heat flux do not vary with depth. Both 
the thermal conductivity, $\kappa$, and the temperature gradient, $\grad T_0$, are constant,
such that $\bm{F}_{\text{cond,0}} = -\kappa \grad T_0 = \text{constant}$.
Under these assumptions, satisfying hydrostatic equilibrium produces an atmosphere defined by
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(z_0 - z)^m \\
T_0(z)    &= T_{t}(z_0 - z).
\label{eqn:polytrope}
\end{split}
\end{equation}
The polytropic index is $m = m_{ad} - \epsilon$
where $m_{ad} \equiv (\gamma-1)^{-1}$ is the adiabatic polytropic index and $\epsilon$ is the
superadiabatic excess.
The subsequent entropy gradient at the top of the atmosphere is $\grad S(L_z) = -\epsilon$. 
Thermodynamic variables are nondimensionalized at the top of the atmosphere as 
$P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, requiring $z_0 \equiv L_z + 1$ and $R = T_{t} = \rho_{t} = 1$.
By this choice, the non-dimensional length scale is the inverse temperature gradient scale and
the timescale is the isothermal sound crossing time of this unit length.
$z$ increases upwards within $[0, L_{z}]$, where $L_{z} = e^{n_{\rho}/m} - 1$ is
determined by the number of density scale heights the atmosphere spans, $n_\rho$.
Throughout this letter, we set $n_{\rho} = 3$ such that the density varies by a factor of 20.
The characteristic timescale of convective dynamics
is related to the atmospheric buoyancy time, $t_{\text{b}} = \sqrt{L_z/g\epsilon}$, with $g = (m+1)$.
We use buoyancy time units in this letter.
